---
title: "rolog: Prolog queries from R"
author: "Matthias Gondan (Department of Psychology, University of Innsbruck, Austria)"
date: "2022-04-21"
bibliography: bibliography.bibtex
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rolog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(rmarkdown.html_vignette.check_title=FALSE)
library(DiagrammeR)
library(DiagrammeRsvg)
library(htmltools)
```

  Matthias Gondan\
  Department of Psychology\
  University of Innsbruck\
  Innrain 9\
  A-6020 Innsbruck\
  Matthias.Gondan-Rochon@uibk.ac.at
  
# Abstract

Prolog is a classical logic programming language with many applications in
expert systems, computer linguistics and traditional, that is, symbolic 
artificial intelligence. The main strength of Prolog is its concise 
representation of facts and rules for the representation of knowledge and
grammar, as well as its very efficient built in search engine for closed world
domains. R is a statistical programming language for data analysis and
statistical modeling which is widely used in academia and industry. Besides the
core library, a lot of packages have been developed for all kinds of statistical
problems, including new-style artificial intelligence tools such as neural
networks for machine learning and deep learning. Whereas Prolog is weak in
statistical computation, but strong in symbolic manipulation, the converse may
be said for the R language. SWI-Prolog is a widely used Prolog system that
offers a wide range of extensions for real world applications, and there already
exist two so-called Prolog packs to invoke R (rserve-client, Real) from
SWI-Prolog. However, given the large user community of R, there may also be a
need for a connection in the reverse direction that allows invoking Prolog
queries in R computations. The R package Rolog embeds the SWI-Prolog system into
an R package, thus enabling deterministic and non-deterministic queries to the
Prolog interpreter. Usage of the Rolog library is illustrated by a few examples.

## Keywords

Statistics; Logic Programming; Artificial Intelligence

# 1. Rolog: Prolog queries from R

The R [@R] programming language and environment is a widely used open source
software for statistical data analysis. The basic R is a functional language 
with lots of support for storage and manipulation of different data types,
and a strong emphasis on operations involving vectors and arrays. Moreover, a
huge number of R packages (e.g., CRAN, https://cran.r-project.org/) have been
contributed that cover problems from diverse areas such as bioinformatics,
machine learning, specialized statistical methods, web programming and
connections to other programming languages. To my knowledge, an interface to
Prolog is lacking so far.

The logic programming language Prolog was invented in the 1970ies by 
Colmerauer [@Colmerauer1996], mostly for the purpose of natural language 
processing. Since then, logic programming has become an important driving force
in research on artificial intelligence, natural language processing, program
analysis, knowledge representation and theorem
proving [@Shoham1994;@Lally2011;@Carro2004;@Hsiang1987]. 
SWI-Prolog [@Wielemaker2012] is an open-source implementation of Prolog that
mainly targets developers of applications, with many users in academia, research
and industry. SWI-Prolog includes a large number of libraries for "the real
world", for example, a web server, encryption, interfaces to C/C++ and other
programming languages, as well as a development environment and debugger. In 
addition, pluggable extensions (so-called packs) are available for specific
tasks to enhance its capabilities.

Unlike R, Prolog is a declarative programming language consisting of facts and
rules that define relations, for example, in a problem space. Prolog's major
strength is its builtin query-driven search engine that efficiently deals with
complex structured data, but this data is not necessarily numerical. In fact,
Prolog only provides a basic collection of arithmetic calculations via a purely
functional interface `is/2`. More complex calculations such as matrix algebra,
statistical models or machine learning need help from other systems, for
example, from R.

Angelopoulos et al. [@Angelopoulos2013] summarize work at the intersection of
symbolic knowledge representation and statistical interence, especially in the
area of model fits [EM algoritms, MCMC, @Sato2001;@Angelopoulos2008] and
stochastic logic programs [@Cussens2000;@Kimmig2011]. One of the major strengths
of logic programming is handling constraints; and a number of systems for
constraint satisfaction tools have been developed (constraint logic programming
on booleans, finite domains, reals, and intervals) for that 
purpose [e.g., @Fruehwirth1998;@Triska2018]. Some constraint handlers exist in
R (see the CRAN task view for optimization problems), but more of them would be
available via a bridge between R and Prolog. 

Earlier approaches to connect Prolog and R have been published as SWI-Prolog 
packs [r..eal, Rserve_client, @Angelopoulos2013;@Rserve] and as a YAP 
module [YapR, @YapR]. Whereas r..eal establishes a direct link to an embedded
instance of R, Rserve_client communicates with a local or remote R 
service [@Urbanek2021]. The former approach emphasises speed, the latter might
be preferred from a security perspective, especially in systems such as 
SWISH [@SWISH] that accept only a set of sandboxed commands for Prolog, but do
not impose restrictions on R. A common feature of the two packages is that they
provide an interface for R calls from Prolog, but not the other way round, that
is, querying Prolog from R is not possible so far.

Rolog is an attempt to fill this gap, and to offer the possibilty to raise
Prolog queries in R scripts, for example, to perform efficient symbolic
computations, searches in complex graphs, parsing natural language and definite
clause grammars. In addition, two Prolog predicates are provided that enable
Prolog to ring back to the R system for bidirectional communication. Similar
to r..real, tight communication between the two systems is established by
linking to a shared library that embeds the current version of SWI-Prolog. The
exchange of data is facilitated by the C++ interfaces of the two
languages [@Edelbuettel2018;@Wielemaker2021]. A less tight connection might be
established using the recently developed machine query interface [@Zinda2021]
that allows socket-based communication between foreign languages and 
SWI-Prolog (and, in fact, the MQI documentation includes an R call).

A bidirectional bridge between R and Prolog might overcome the limitations of 
both languages, thereby combining the extensive numerical and statistical power
of the R system with Prolog's skills in the representation of knowledge and
reasoning. In addition to the useful little tools shown in the examples below, 
Rolog can therefore contribute to progress at the intersection of traditional
artificial intelligence and contemporary statistical programming.

The next section presents the interface of Rolog in detail. Section\ 3 presents
possible extensions of the package at both ends, in R and Prolog. Section\ 4 is
a list of illustrative examples that offer useful extensions to the R system.
Conclusions and further perspectives are summarized in Section 5.

# 2. Basic syntax

Rolog has a rather minimalistic syntax, providing only some basic ingredients
to establish communication with an embedded SWI-Prolog. Some ways to extend the 
interface are described in Section\ 3.

After installation (for now, with `remotes::install_github("mgondan/rolog")`,
later, with `install.packages("rolog")`), the package is loaded in the standard
way using R's `library`-command.

```{r}
library(rolog)
```

We can see SWI-Prolog's typical welcome message.

## R interface

Most of the work can done using the three R functions `query`, `submit`,
and `clear`. The functions `consult`, `once`, and `findall` are provided for
convenience.

*consult*. In most applications, a number of Prolog facts and rules will be 
loaded into the system. To facilitate this recurrent task, the prolog
directive `consult/1` has been mirrored into R, `consult(filename)`, 
with _filename_ given as a string (or a list of strings if multiple files are to
be consulted). Note that the full filename should be given, including the
extension (e.g., ".pl"). The function returns `TRUE` on success, 
otherwise `FALSE` and an error message is shown.

*query*. The function `query(call, options)` is used to create a Prolog 
query (without invoking it yet). The first argument _call_ is a regular R call
that is created using R's function `call(name, ...)`. This call represents the
Prolog predicate which will be queried in the later course. The creation of such
predicates and Prolog terms is described below and can become quite 
cumbersome (see the examples in Section\ 4). The second argument, _options_, may
be used for ad hoc modifications of the translation between R and Prolog, see
the section below. The function returns `TRUE` on success. Note that the
function does not check if the corresponding Prolog predicate exists (but see
`submit()` below).

Only a single query can be opened at a given time. If a new query is created
while another query is still open, a warning is shown and the other query is
closed.

*submit*. Once a query has been created, it can be submitted using `submit()`.
If the query fails, the return value is `FALSE`. If the query succeeds, a list
of constraints is returned, with bindings for the variables that satisfy the
query. Repeated calls to submit are possible, returning the different solutions
of a query (until it eventually fails). Programmatically distinguishing between
the different types of return values for success and failure (list vs. `FALSE`)
is facilitated by the R function `isFALSE(x)`.

*clear*. Closes the query. The name of the function is chosen to avoid name
clashes with R's own built-in function close. The function returns an invisible
`TRUE`, even if there is no open query.

The following short R program illustrates a query to Prolog's `member/2` using 
Rolog's syntax rules.

```{r}
# member(1, [1, 2.0, a, "b", X])
query(call("member", 1L, list(1L, 2.0, quote(a), "b", expression(X), TRUE)))

# returns an empty list, stating that the query satisfied
submit()

# returns a list, stating that the query is satisfied if X = 1
submit()

# close the query
clear()
```

Listing\ 1.
: A query to Prolog's `member/2` predicate.

*once* and *findall*. The function `once(call, options)` is a convenience
function that acts as a shortcut for `query(call, options)`, `submit()`, 
and `clear()`. Similarly, `findall(call, options)` abbreviates the 
commands `query(call, options)`, repetition of `submit()` until failure,
and `clear()`, returning a list collecting the the return value of the 
individual calls to submit.

## Creating Prolog terms in R

Table\ 1 summarizes the rules for the translation from R to Prolog. Most rules
work in both directions, but a few exceptions exist. For example, there is an
empty atom in Prolog, but no empty symbol in R, so the empty atom is translated
to a character string in R. 

Table\ 1
: Creating Prolog terms from R

|R                       |Prolog                   |Note/Alternatives        |
|:-----------------------|:------------------------|:------------------------|
|`expression(X)`         |Variable X               |not necessarily uppercase|
|`as.symbol(abc)`        |Atom abc                 |`as.name`, `quote`       |
|`TRUE`, `FALSE`, `NULL` |Atoms true, false, null  |                         |
|`"abc"`                 |String \"abc\"           |                         |
|`3L`                    |Integer 3                |                         |
|`3`                     |Float 3.0                |                         |
|`call("term", 1L, 2L)`  |term(1, 2)               |                         |
|`list(1L, 2L, 3L)`      |List [1, 2, 3]           |                         |
|`list(a=1, b=2, c=3)`   |List [a-1, b-2, c-3]     |                         |
|`c(1, 2, 3)`            |#(1.0, 2.0, 3.0)         |vectors of length > 1    |
|`c(1L, 2L, 3L)` or `1:3`|\'%\'(1, 2, 3)           |                         |
|`c("a", "b", "c")`      |\$\$(\"a\", \"b\", \"c\")|                         |
|`c(TRUE, FALSE, NA)`    |!(true, false, na)       |                         |

Moreover, R is mostly vectorized, lacking support for scalar entities, that is,
scalar entities are treated as vectors of length\ 1. Conversely, Prolog does not
natively support vectors or matrices. The problem is solved in the following
way: 

* R vectors of length 0 are translated to Prolog's empty list.
* R vectors of length 1 are translated to Prolog scalars.
* R vectors of length _N_ > 1 are translated to Prolog 
  terms `#/N`, `%/N`, `$$/N`, and `!/N` for floating points, integers, strings
  and logicals, respectively.

In the reverse direction, Prolog terms like `#/N` are translated back to R
vectors of length _N_. This including the terms `#/0` and `#/1` that normally
don't occur. To summarize, the rules for translation are not fully symmetrical.
A quick check for symmetry of the representation is obtained by a query to
`r_eval/2` (see also below, subsection Prolog interface):

```{r}
once(call("r_eval", c(1, 2, NA, NaN, Inf), expression(X)))
```

## Rolog options

A few package-specific options have been defined to allow some finetuning of
the rules for translation between R and Prolog.

* realvec (string): Name of the prolog term for vectors of floats (default is #)
* intvec (string): same for vectors of integers (default is %)
* boolvec (string): same for vectors of logicals (default is !)
* charvec (string): same for vectors of character strings (default is \$\$). The
  single dollar cannot be used because it is the list operator in R.
* scalar (logical): if `TRUE` (default), R vectors of length 1 are translated to
  scalars in Prolog. If `FALSE`, R vectors are always translated to `#/N` etc.,
  depending on the type.
* portray (logical): if `TRUE (default)`, the result of `query`, `once` and 
  `findall` includes an attribute with a representation of the query in Prolog.

The command `rolog_options()` returns a list with all the options. The options
can be globally modified like e.g. this:

```r
options(rolog.intvec="%%")
```

In a given query, the options can be set in the optional argument, e.g.

```r
query(call("member", expression(X), list(1:3, 4:6)), list(intvec="%%"))
```

## Prolog interface

Rolog offers some basic support to call R from Prolog, that is, connecting the
two systems in the reverse direction. Two predicates can be used for this 
purpose, `r_eval(Call)` and `r_eval(Function, Result)`. The former just invokes
R with the command _Call_ (ignoring the result); the latter evaluates _Function_
and unifies the result with _Result_. Note that proper quoting of R functions
is needed at the Prolog end, especially with R functions that start with
uppercase letters or contain a dot in their name.

Two use cases for `r_eval/2` are shown in Section\ 4 below.

# 3. Extending the package

The package is intentionally kept minimalistic, but can easily be extended by
convenience functions on both ends, Prolog and R, to facilitate recurrent tasks
and/or avoid cumbersome syntax. R is a functional language, whereas Prolog is
declarative. Obviously, there cannot be a perfect one-to-one correspondence
between the syntactic components of two programming languages that follow
completely different paradigms. Whereas symbols, functions, numbers and
character strings are easily mapped between R and Prolog, there are loose ends 
at both sides. In particular, Prolog variables are translated from and to 
R _expressions_ (not to be confused with R symbols), and R vectors of 
length _N_ > 1 are translated to the Prolog terms `#/N`, `%/N`, `!/N`,
and `$$/N`, as mentioned above.

These rules are, in principle, arbitrary and can be intercepted at several 
stages of the process:

* R functions that may be used to preprocess specific R elements before
  translation into Prolog (see, e.g., the R function `as.rolog()`)
* Prolog wrappers that manipulate the term before it is called and afterwards
  (see the dicts below)
* R functions that postprocess the result of a query

The process is illustrated in Figure\ 1.

```{r, echo=FALSE, fig.width=6, fig.height=2}
HTML(export_svg(grViz(
  'digraph G
   {
     rankdir=LR
     Query Result

     subgraph cluster_0 
     {
       style=filled
       color=lightgrey
       node [style=filled,color=white]
       r2rolog -> forth -> rolog_pl
     }

     subgraph cluster_1 
     {
       style=filled
       color=lightgrey
       node [style=filled,color=white]
       rolog2r -> back [dir=back]
       back -> pl_rolog [dir=back]
     }
  
     Query -> r2rolog
     rolog_pl:e -> Prolog
     pl_rolog:e -> Prolog [dir=back]
     Result -> rolog2r [dir=back]

     Query [shape=Mdiamond;width=0.7;height=0.7]
     r2rolog [shape=rect,label="preproc(...)"]
     forth [label="(rolog)"]
     rolog_pl [shape=rect,label="preproc/2"]
     Prolog [shape=Mcircle]
     pl_rolog [shape=rect,label="postproc/2"]
     rolog2r [shape=rect,label="postproc(...)"]
     back [label="(rolog)"]
     Result [shape=Msquare]
   }')))
```

Figure\ 1
: Workflow in Rolog

## Preprocessing in R

We have already seen above that raising even simple everyday Prolog queries such
as `member(X, [1, 2, 3, a, b])` require complicated R expressions such as

```r
query(call("member", expression(X), list(1, 2, 3, quote("a"), quote(b)))
```

The R function `as.rolog(call)` is meant to simplify this a bit by
translating symbols starting with a dot to Prolog variables, and calls
like `""[1, 2, 3, a, b]` to lists. The argument _call_ is typically 
a quoted R call or symbol:

```r
q <- quote(member(.X, ""[1, 2, 3, a, b]))
query(as.rolog(q))
```

Note that the name of the variable will still be _X_ in the later course, 
not "dot-X". A bit flexibility is lost because `quote()` treats the
arguments _a_, _b_ as symbols; to evaluate them (i.e., "unquote"), they can be
put in parentheses:

```r
a <- 4
b <- 5
q <- quote(member(.X, ""[1, 2, 3, a, (b)]))
query(as.rolog(q))
```

resulting in the query `member(X, [1, 2, 3, a, 5])`. More sophisticated work
with quasi-quotations (i.e., "unquoting" expressions) is described in 
the "Advanced R" book [@Wickham2019].

Section\ 4 includes an example for mathematical rendering of R expressions. In
that example, a preprocessing function is used to bring function calls with
named arguments to a canonical form which is then handled in Prolog.

## Postprocessing in R

This may again be a function that reverts some of the manipulations during
preprocessing. For `once()` and `submit()`, such a function would operate on the
bindings. For example, many Prolog programmers are used to operate with atoms
instead of character strings, which is the preferred representation of symbolic
information in R. The following simple example illustrates conversion of the
results for a query like `member(X, [a, b, c])` to strings.

```{r}
stringify <- function(x)
{
  # replace Prolog variable by the value of an R variable with the same name
	if(is.name(x))
	  return(as.character(x))
	
	# Recurse into lists and calls
	if(is.call(x))
	  x[-1] <- lapply(x[-1], FUN=stringify)

	if(is.list(x))
	  x <- lapply(x, FUN=stringify)

  # Leave the rest unchanged
	return(x)
}

# This may be the return value of a Prolog query
q <- quote(member(.X, ""[a, b, c]))
r <- findall(as.rolog(q))
stringify(r)
```

## Pre- and postprocessing in Prolog

Recent versions of SWI-Prolog support so-called dictionaries of the 
form `Tag{Key1:Value1, Key2:Value2, ...}`. The tag is typically an 
atom (but can be a variable, as well), the keys are unique atom or integers;
the values can be anything. Suppose we have a Prolog predicate that does
something with dicts, and we would like to query it from R. The simplest
solution is a wrapper in Prolog that translates key-value 
pairs `[Key1-Value1, Key2-Value2, ...]` back and forth to dicts:

```prolog
do_something_with_pairs(Pairs0, Pairs1) :-
    dict_pairs(Dict0, my_dict, Pairs0),
    do_something_with_dicts(Dict0, Dict1),
    dict_pairs(Dict1, my_dict, Pairs1).
```

`do_something_with_pairs/2` can then be queried from R using, e.g., lists with
named elements.

```r
once(call("do_something_with_pairs", list(a=1, b=2), expression(X)))
```

In the code above, `dict_pairs/2` takes the role of both `preproc/2`
and `postproc/2` in Figure\ 1. It illustrates that complicated syntax on the R
side when can be much simpler to do the conversion at the Prolog end.

A way to extend Prolog by add-ons ("packs") are shown in the next section.

# 4. Examples

In this section I present a few usage examples for the Rolog package in
increasing complexity. Although the code snippets are mostly self-explanatory,
some familiarity with the Prolog language is helpful.

## Hello, world

Prolog's typical _hello world_ example is a search through a directed acyclic
graph (DAG), for example, a family tree like the one given in Listing\ 1.

```prolog
parent(pam, bob).
parent(tom, bob).
parent(bob, ann).
parent(bob, pat).
parent(pat, jim).

ancestor(X, Z) :-
    parent(X, Z).

ancestor(X, Z) :-
    parent(X, Y),
    ancestor(Y, Z).
```

Listing\ 1
: A family tree in Prolog (see also family.pl)

Listing\ 1 is included in the package and can be accessed from R using the
function `system.file(...)`. Within Prolog, the normal workflow is to consult the
code with `[family]` and then to raise queries such as `ancestor(X, jim)`, which
returns, one by one, four solutions for the variable _X_. In R, we obtain the
following results:

```{r}
library(rolog)

# [family].
consult(system.file(file.path("pl", "family.pl"), package="rolog"))

# ancestor(X, jim).
query(call("ancestor", expression(X), quote(jim)))

# solutions for X, one by one
submit()
submit()
submit()
submit()
submit() # no more results (closing the query)
submit() # warning that no query is open
# clear() # normally used to close a query
```

As stated above, `consult()` loads the facts and rules of Listing\ 1 into the
database. `query(expr)` initializes the query `expr`, and the subsequent calls
to `submit` return the conditions under which the query succeeds. In this 
example, the query succeeds if _X_ is either `pat`, `pam`, `tom` or `bob`. 
A query is closed with `clear()`, or automatically if the query fails. Note that
it is generally not possible to open two queries simultaneously, so opening a
second query while another one is still open will raise a warning. If we are
interested in just the first solution, we can use `once(expr)` as a shortcut
to `query(expr)`-`submit()`-`clear()`. If we want to collect all solutions of a
query with a finite set of solutions, we can use `findall(expr)`.

A rather cumbersome aspect of Rolog is the construction of Prolog
terms and queries. `expression(X)` encapsulates the variable `X`, R symbols from
`quote(jim)` or `as.symbol("jim")` are translated to Prolog atoms, and Prolog
compounds such as `ancestor/2` are generated using `call("ancestor", ...)`. 
A simplified syntax is provided by `as.rolog(...)` that accepts quoted
expressions with dots indicating Prolog variables:

```r
q = quote(ancestor(.X, jim))
findall(as.rolog(q))
```

Note that `as.rolog(...)` removes the dot from the variable name.

## Backdoor test

A more useful application of DAGs is confounder adjustment in causal 
analysis [@greenland1999;@ggdag]. The Prolog file `backdoor.pl` is an 
implementation of Greenland et al.'s criteria for the backdoor test 
for _d_-separation in DAGs, with a predicate `minimal/3` that searches for
minimally sufficient sets of variables for confounder adjustment on the causal
path between exposure and outcome.

```{r}
# [backdoor].
consult(system.file(file.path("pl", "backdoor.pl"), package="rolog"))

# Figure 12 in Greenland et al.
add_node = function(N)
	invisible(once(call("assert", call("node", N))))

add_arrow = function(X, Y)
	invisible(once(call("assert", call("arrow", X, Y))))

add_node("a")
add_node("b")
add_node("c")
add_node("d") # outcome
add_node("e") # exposure
add_node("f")
add_node("u")

add_arrow("a", "d")
add_arrow("a", "f")
add_arrow("b", "d")
add_arrow("b", "f")
add_arrow("c", "d")
add_arrow("c", "f")
add_arrow("e", "d")
add_arrow("f", "e")
add_arrow("u", "a")
add_arrow("u", "b")
add_arrow("u", "c")

findall(call("minimal", "e", "d", expression(S)))
```

The query to `minimal/3` returns two minimally sufficient sets of covariates for
confounder adjustment (namely, {a, b, c} and {f}).

## Definite clause grammars

One of the main driving forces of Prolog development was natural language 
processing [@Dahl1981]. Therefore, the next example is an illustration of 
sentence parsing using so-called definite clause grammars. As Listing\ 3 shows,
Rolog can access modules from SWI's standard library (here, it 
is "dcg/basics.pl").

```prolog
:- use_module(library(dcg/basics)).

% Translate R string to code points and invoke phrase/2
sentence(Tree, Sentence) :-
    string_codes(Sentence, Codes),
    phrase(s(Tree), Codes).

% Simple grammar with sentences, noun, verb and participle phrases
s(s(NP, VP)) --> np(NP, C), blank, vp(VP, C).
np(NP, C) --> pn(NP, C).
np(np(Det, N), C) --> det(Det, C), blank, n(N, C).
np(np(Det, N, PP), C) --> det(Det, C), blank, n(N, C), blank, pp(PP).
vp(vp(V, NP), C) --> v(V, C), blank, np(NP, _).
vp(vp(V, NP, PP), C) --> v(V, C), blank, np(NP, _), blank, pp(PP).
pp(pp(P, NP)) --> p(P), blank, np(NP, _).

% Determiners, personal nouns, nouns, verbs and participles
det(det(a), sg) --> `a`.
det(det(the), _) --> `the`.
pn(pn(john), sg) --> `john`.
n(n(man), sg) --> `man`.
n(n(men), pl) --> `men`.
n(n(telescope), sg) --> `telescope`.
v(v(sees), sg) --> `sees`.
v(v(see), pl) --> `see`.
v(v(saw), _) --> `saw`.
p(p(with)) --> `with`.
```

Listing\ 2
: Simple grammar and lexicon. `sentence/2` preprocesses the R call.

As in the first example, we first consult a little Prolog program with a
minimalistic grammar and lexicon (Listing\ 2), and then raise a query asking for
the syntactic structure of "john saw a man with a telescope". Closer inspection
of the two results reveals the two possible meanings, "john saw a man _who
carries_ a telescope" versus "john saw a man _through_ a telescope". More Prolog
examples of natural language processing are found 
in Blackburn and Bos [@Blackburn2005], including the resolution of anaphoric
references and the extraction of semantic meaning.

```{r}
# [telescope].
consult(system.file(file.path("pl", "telescope.pl"), package="rolog"))

# findall(sentence(Tree, "john saw a man with a telescope")).
findall(call("sentence", expression(Tree), "john saw a man with a telescope"))
```

## Installation of add-ons for Prolog

In description of the previous example, I noted in passing that Rolog can use
the built-in libraries of SWI-Prolog (e.g., by calls to `use_module/1`). It is
also possible to extend the installation by add-ons, including add-ons that
require compilation, if the build tools (essentially, RTools) are properly
configured. This is illustrated below by the demo add-on "environ" [@Environ]
that collects the current environment variables.

```r
# pack_install(environ, [interactive(false)]).
once(call("pack_install", as.name("environ"), list(call("interactive", FALSE))))

# use_module(library(environ))
once(call("use_module", call("library", quote(environ))))

# environ(X)
once(call("environ", expression(X)))
```

The query then unifies _X_ with a list with Key=Value terms. The purpose if this
example is obviously not to mimick the built-in function `Sys.getenv()` from R,
but to illustrate the installation and usage of Prolog extensions from within R.

## Term manipulation

Prolog is homoiconic, that is, code is data. In this example, we make use of
Prolog's ability to match expressions against given patterns and modify these
expressions according to a few predefined "buggy rules" [@Brown1978], inspired
by recurrent mistakes in the statistics exams of my students. Consider 
the $t$-statistic for comparing an observed group average to a population mean:

$$
T = \frac{\overline{X} - \mu}{s / \sqrt{N}}
$$

Some mistakes may occur in this calculation, for example, omission of the
implicit parentheses around the numerator and the denominator when typing the
numbers into a calculator, resulting
to $\overline{X} - \frac{\mu}{s} \div \sqrt{N}$, or forgetting the square root
around $N$, or both. Prolog code for the two buggy rules is given in Listing\ 3.

```Prolog
% Correct step from task to solution
expert(tratio(X, Mu, S, N), frac(X - Mu, S / sqrt(N))).

% Mistakes
buggy(frac(X - Mu, S / SQRTN), X - frac(Mu, S) / SQRTN).
buggy(sqrt(N), N).

% Apply expert and buggy rules
step(X, Y) :-
    expert(X, Y).

step(X, Y) :-
    buggy(X, Y).

% Enter expressions
step(X, Y) :-
    compound(X),
    mapargs(search, X, Y),
    dif(X, Y).

% Search through problem space
search(X, X).

search(X, Z) :-
    step(X, Y),
    search(Y, Z).
```

Listing 3
: Manipulating terms in Prolog

The little e-learning system shown in Listing\ 2 (`buggy.pl`) is invoked using
the R script below.

```{r}
library(rolog)
consult(system.file(file.path("pl", "buggy.pl"), package="rolog"))

q <- quote(search(tratio(x, mu, s, n), .S))
findall(as.rolog(q))
```

The fourth and the sixth result are combinations of the two buggy 
rules (parenthesis, then square root, and the other way round). Some additional
filters would be needed to eliminate trivial and redundant solutions [see, e.g.,
the chapter on generate-and-test in @Sterling1994].

It should be mentioned that R is homoiconic, too, and the Prolog code above can,
in principle, be rewritten in R using non-standard evaluation 
techniques [@Wickham2019]. Prolog's inbuilt pattern matching algorithm 
simplifies things a lot, though. An important feature of such a term 
manipulation is that the evaluation of the term can be postponed; for example, 
there is no need to instantiate the variables `X`, `Mu`, `s` and `N` with
given values before raising a query. This is especially helpful for variables 
that may represent larger sets of data in the later course.

## Rendering mathematical expressions

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
However, so far, R expressions such as `pbinom(k, N, p)` are typeset as-is; 
prettier mathematical expressions such as $P_\mathrm{Bi}(X \le k; N, p)$ require
Latex commands like `P_\mathrm{Bi}\left(X \le k; N, p\right)`, which are
cumbersome to type in and hardly readable even for simple expressions. Below we
make use of Prolog's grammar rules for the automatic translation of R
expressions to MathML. The result can then be used for calculations or it can be
rendered on a web page. A limited set of rules for translation from R to MathML
is found in `pl/mathml.pl` of the Rolog package. The relevant code snippets are
shown in the listings below, along with their output.

```{r}
library(rolog)
consult(system.file(file.path("pl", "mathml.pl"), package="rolog"))

# R interface to Prolog predicate r2mathml/2
mathml = function(term)
{
  t = once(call("r2mathml", term, expression(X)))
  cat(paste(t$X, collapse=""))
}
```

Listing 4
: Using Rolog to generate MathML from R expressions

The first example is easy. At the Prolog end, there is a handler for `pbinom/3`
that translates the term into a pretty MathML syntax like P_bi(X <= k; N, pi).

```{r, results="asis"}
term = quote(pbinom(k, N, p))

# Pretty print
mathml(term)

# Do some calculations with the same term
k = 10
N = 22
p = 0.4
eval(term)
```

The next example is interesting because Prolog needs find out the name of
the integration variable for `sin`. For that purpose, Rolog provides a 
predicate `r_eval/2` that calls R from Prolog (i.e., the reverse direction, see
also next example). Here, the predicate is used for the 
R\ function `formalArgs(args(sin))`, which returns the name of the function
argument of `sin`, that is, `x`.

```{r, results="asis"}
term = quote(integrate(sin, 0L, 2L*pi))
mathml(term)
eval(term)
```

Note that the Prolog end, the handler for `integrate/3` is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, `integrate(sin, lower=0L, upper=2L * pi)` would not print the desired
result.

Therefore, `pl/mathml.pl` includes two handlers that accept terms with named
arguments, `integrate(f=Fn, lower=Lower, upper=Upper)`, as well as terms of the
form `$(integrate(Fn, Lower, Upper), value)` that are needed for the evaluation
below.

```{r, results='asis'}
# Apply match.call to all components of a term
canonical <- function(term)
{
  if(is.call(term))
  {
    f <- match.fun(term[[1]])
    if(!is.primitive(f))
      term <- match.call(f, term)
    
    # Recurse into arguments
    term[-1] <- lapply(term[-1], canonical)
  }

  return(term)
}

# A custom function
g <- function(u)
{
  sin(u)
}

# Mixture of (partially) named and positional arguments in unusual order
term <- quote(2L * integrate(low=-Inf, up=Inf, g)$value)
mathml(canonical(term))

# It is a bit of a mystery that R knows the result of this integral.
eval(term)
```

The extra R function `canonical()` applies `match.call()` to non-primitive R
calls, basically cleaning up the arguments and bringing them into the correct
order.

## Calling Prolog from R

The basic workflow of the bridge from R to Prolog is to (A)\ translate an 
R\ expression into a Prolog term (i.e., a predicate), (B)\ query the predicate,
and then, (C)\ translate the result (i.e., the bindings of the variables) back
to R. The reverse direction is straightforward, we start by translating a
Prolog term to an R expression (i.e. Step\ C), evaluate the R expression, and 
then translate the result back to a Prolog term (Step A). Rolog provides two
Prolog predicates for this purpose, `r_eval(Expr)` and `r_eval(Expr, Res)`. The
former is used to invoke an R expression *Expr* for its side effects (e.g.,
initializing a random number generator); it does not return a result. The latter
is used to evaluate the R expression *Expr* and return the result *Res*. The
code snippet in Listing\ 5 (`r_eval.pl`) illustrates this behavior.

```prolog
r_seed(Seed) :-
    r_eval('set.seed'(Seed)).

r_norm(N, L) :-
    r_eval(rnorm(N), L).
```

```{r}
# [r_eval].
consult(system.file(file.path("pl", "r_eval.pl"), package="rolog"))

# rnorm(3)
once(call("r_norm", 3L, expression(X)))
```

Listing\ 5
: Calling Prolog from R using `r_eval/1` and `r_eval/2`. The R\ call `set.seed` is quoted because the dot is an operator in Prolog.

The example in Listing\ 5 is a bit trivial, basically illustrating the syntax
and the workflow. More serious applications of `r_eval/1,2` are illustrated in
the example on mathematical expressions where `r_eval/2` is used to obtain
a names of a function argument, as well as in the next example on interval
arithmetic, where `r_eval/2` is used with monototonically behaving R functions.

## Interval arithmetic

Let $\langle\ell, u\rangle$ denote a number between $\ell$ and $u$, $\ell\le u$.
It is easily verified that the result of the
difference $\langle\ell_1, u_1\rangle - \langle\ell_2, u_2\rangle$ is somewhere
in the interval $\langle \ell_1 - u_2, u_1 - \ell_2\rangle$, and a number of
rules exist for basic arithmetic operations and (piecewise) monotonically
behaving functions [@Hickey2001]. For ratios, denominators with mixed sign yield
two possible intervals, for example,
$\langle 1, 2\rangle / \langle -3, 3\rangle = \langle -\infty, 3\rangle \cup \langle 3, \infty\rangle$,
as shown in Figure 4 in Hickey et al.'s article. The number of possible
candidates increases if more complicated functions are involved, as unions of
intervals themselves appear as arguments (e.g., if $I_1 \cup I_2$ is added
to $I_3 \cup I_4$, the result
is $I_1 + I_3 \cup I_1 + I_4 \cup I_2 + I_3 \cup I_2 + I_4$). As a consequence,
calculations in interval arithmetic are non-deterministic in nature, and the
number of possible results is not foreseeable and cannot, in general, be
vectorized as is often done in R. Use cases for interval arithmetic are the
limitations of floating-point representations in computer hardware, but
intervals can also be used to represent the result of measurements with limited
precision, or truncated intermediate results of students doing hand
calculations. A few rules for basic interval arithmetic are found
in `pl/interval.pl`; a few examples are shown below. Again, Prolog rings back to
R via `r_eval/2` to determine the result of `dbinom(X, Size, Prob, Log)`.

```{r}
# [interval].
consult(system.file(file.path("pl", "interval.pl"), package="rolog"))

# findall(1 ... 2 / -3 ... 3, Res).
q <- quote(int(`...`(1, 2) / `...`(-3, 3), .Res))
findall(as.rolog(q))

# t-ratio
D  = quote(`...`(5.7, 5.8))
mu = 4
s  = quote(`...`(3.8, 3.9))
N  = 24L
tratio = call("/", call("-", D, mu), call("/", s, call("sqrt", N)))
findall(call("int", tratio, expression(Res)))

# Binomial density
prob = quote(`...`(0.2, 0.3))
once(call("int", call("dbinom", 4L, 10L, prob, FALSE), expression(Res)))

prob = quote(`...`(0.5, 0.6))
once(call("int", call("dbinom", 4L, 10L, prob, FALSE), expression(Res)))

prob = quote(`...`(0.2, 0.6))
once(call("int", call("dbinom", 4L, 10L, prob, FALSE), expression(Res)))
```

The slightly cumbersome syntax for entering an interval $\langle \ell, u\rangle$
is due to the fact that the ellipsis is a reserved symbol in R and cannot be
used as an infix operator. A way more powerful and comprehensive system for
constraint logic programming over intervals is available as a Prolog 
pack [@Workman2021] and can easily be connected to R using, for example, the
present package.

# 5. Conclusions

R has become the primary language for statistical programming and data science,
but is currently lacking support for traditional, symbolic artificial intelligence.
There are already two addons for SWI-Prolog that allow to run R calculations
from Prolog [@Angelopoulos2013;@Rserve], but a connection in the other direction
was missing, so far. Rolog bridges this gap by providing an interface to a
SWI-Prolog distribution embedded into an R package. The communication between
the two systems is mainly in the form of queries from R to Prolog, but two
Prolog predicates allow Prolog to ring back and evaluate terms in R. The design
of the package is minimalistic, providing three main 
functions `query(...)`, `submit()`, and `clear()`, and a very limited set of
convenience tools (`consult(...)`, `once(...)`, and `findall(...)`) to
facilitate everyday actions. As both systems are homoiconic in nature, it was easy to
establish a one-to-one correspondence between the elements of the two languages. Most
exceptions (e.g., lack of R support for empty symbols) can be avoided and/or 
circumvented by wrapper functions at both ends.

Simple ways to extend the package have been described in Section\ 3; such
extensions could, for example, include R objects and structures like those
returned by `lm`, or S4 classes. In many use cases, this may be realized by
transforming the R object to a list with named elements, and "reconstruct" the
object on the Prolog end if needed. After a query, the process is reversed.
If speed is an issue, more of these steps can, in principle, be moved into the
package and implemented in Rcpp.

Rolog, thus, opens up a wide of applications in logic programming for
statisticians and researchers at the intersection of symbolic and connectionist
artificial intelligence, where concise knowledge representation is combined with
statistical power. Moreover, Rolog provides starting points for useful
small-scale solutions for everyday issues in data science (term transformations,
pretty mathematical output, interval arithmetic).

At its present stage, a major limitation of Rolog is its rather slow speed. For
example, translation of R lists or vectors to the respective elements of the
Prolog language (lists, `#/N`) is done element-wise, in both directions. I tried
to optimize this process using Rcpp [@Edelbuettel2018], but Prolog does not
support vectors or matrices, so there remains an upper bound in the efficiency.
Of course, Prolog's primary purpose is not vector or matrix calculation, so this
limitation may not show up in real applications. Another issue, maybe a bit 
annoying, is the rather cumbersome syntax of the interface, with the need for
quoted calls and R expressions being a bit misused for representing Prolog
variables. Rolog was deliberately chosen to be minimalistic and, so far, only
depends on base R. A more concise form of representation might be obtained by
tools from the "Tidyverse" ecosystem, as described in Chapter\ 19 of Advanced
R [@Wickham2019]. Finally, at this stage, Rolog is unable to deal with cyclic
Prolog terms (e.g., `once(call("=", expression(A), call("f", expression(A))))`
raises an error message).

The source code of the package is found at https://github.com/mgondan/rolog/,
including installation instructions for Unix, Windows and macOS. At the moment,
Rolog is a source package that requires the RTools 4.2 toolchain under Windows. 
Binaries will be made available soon on https://cran.r-project.org/.

# Acknowledgement

I am grateful to Jan Wielemaker, the main developer of SWI-Prolog; this package
would never have been finalized without his help. Moreover, development of the
package profited substantially from the Prolog packs rserve_client [@Rserve] and 
real [@Angelopoulos2013]. The project was financially supported by the European
Commission (Erasmus+ project QHelp, 2019-1-EE01-KA203-051708).

# References
